---
title: "Preparing Your Computer for SNA"
author: "Douglas Luke"
date: "7/9/2021"
output: pdf_document
urlcolor: blue
---

# Introduction

There are only a few things that need to be done to set
up a computer for the Social Network Analysis workshop. 
Assuming you have a reasonably new computer 
with a fairly modern operating system, these steps should take you 
less than 15 minutes. Specifically, you should do four things
to prepare your computer:

1. Download and install R
2. Download and install RStudio
3. Within RStudio, install two network analysis packages
4. Within RStudio, install the UserNetR dataset package

You may already have R and RStudio installed on your computer as
part of a previous class or work that you do. In that case, you
can skip the first two steps, but make sure you have the most
recent versions of both R and RStudio installed.

# Installing needed software

## Step 1: Download & install R

For Windows machines:
- Go to https://cran.r-project.org/bin/windows/base/
- Click **Download R 4.1.0 for Windows** (current version as of 7.8.21) 
- Open the .exe file and follow recommended install instructions and settings

For Mac OS:
- Go to https://cran.r-project.org/bin/macosx/ 
- Click **R-4.1.0.pkg** (current version as of 7.8.21)
- Install following recommended install instructions and settings

* Run \emph{R} to make sure that it is installed properly.
* You can type in the following commands just to make sure that it is working correctly.


```{r}
dum <- 1:10
dum2 <- dum^2
dum
dum2
```


## Step 2: Download & install RStudio

For all:
- Go to https://www.rstudio.com/products/rstudio/download/#download 
- Click on the installer **RStudio 1.4.1106** (current version as of 4.8.21) appropriate for your OS
- Follow recommended install instructions and settings


# Installing required network packages

## Step 3: Installing `statnet` and `igraph`

The basic functionality of R can be extended by using various packages. These
are created by other statisticians and scientists, and are free to install and use.
For this class we will be using two different network analysis packages: *statnet* and
*igraph*. For the following code (and all coding we will do in this class), this
will be run within the RStudio IDE. If you are new to R/RStudio, just type
each code line into the RStudio console window.

The following code will install the two network packages.

```{r eval=FALSE}
install.packages("statnet")
install.packages("igraph")
```

## Step 4: Installing the *UserNetR* package 

I have created and collected a number of example datasets that accompany 
my Springer book, and these will be
used in this class. They are available as an R package, which can
be downloaded and installed within R itself.

* Start RStudio, and make sure it is working properly.
* Install the *devtools* package by typing in the following line of code in the console
 (by default, the window in the lower left-hand corner):
 
```{r eval=FALSE}
install.packages("devtools")
```

* Assuming this worked with no errors, then type in the following code to 
install the *UserNetR* package from GitHub:

```{r eval=FALSE}
library(devtools)
install_github("DougLuke/UserNetR")
```

# Testing everything

The following code should then work to access one of the datasets, present some
basic information about the network, and plot the network. The network plot itself may look a bit different when you plot.

```{r,message=FALSE,warning=FALSE}
library(statnet)
library(UserNetR)
data(Moreno)
summary(Moreno,print.adj=FALSE)
plot(Moreno, vertex.col = "gender")
```









