---
title: "Lab 1 - The Five-Number Summary"
author: "Douglas Luke"
date: "July 10, 2021"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: false
    theme: lumen
    highlight: textmate
urlcolor: blue
---

## Introduction


This is the first SNA Lab for the summer workshop. (See *Preparing Your Computer for SNA*
for instructions
on how to install R, RStudio, and the required datasets.)

This Lab has two goals:

1. Demonstrate the *Notebook* functionality built-in to RStudio.
2. Demonstrate some very basic ways to examine a network using the
`statnet` package in R.

First, load the required network and data packages.

```{r, message=FALSE}
library(statnet)
library(UserNetR)
data(Moreno)
```

## Simple visualization

```{r}
plot(Moreno,
     vertex.col = (Moreno %v% "gender" + 2),
     vertex.cex=3)
```

E: Try changing the number for 'vertex.cex' and see what happens when you rerun the code.

## Basic Description

When you have network data, there are usually some basic 
characteristics of the network that you will interested in. 

### Network summary for a statnet object

Most things that R can manipulate (plot, analyze, transform, etc.)
are stored as some type of R object. These objects can be
examined in a number of ways. You can usually start by
calling the *summary()* function on the object that you're 
interested in. Here, we can look at the basic information
about the Moreno network. 

```{r}
summary(Moreno,print.adj=FALSE)
```

E: Try removing the 'print.adj = F' part of the command and see if you
can figure out what is happening.

### Network size

The most fundamental piece of information about a network is its
size. This is the number of nodes (vertices) that are in the network.
(Typically, you will already know the size of the network before you
even look at the data.)

```{r}
network.size(Moreno)
```

Notice that this information was also included in the network summary
above.

### Network density

Possibly the most useful basic descriptive of a network is its density,
which is the proportion of observed ties to the total number of possible
ties. It is a measure of the *interconnectedness* of the network.

```{r}
gden(Moreno)
```

### Components

If you examine the previous plot, you can see that the network
has two separate groups. Most of the members are connected to each
other, but two green dots are connected to each other but not to the larger group.
These are called components, and you can see how many are in your
network with the following code.

```{r, message=FALSE,warning=FALSE}
components(Moreno)
```

### Diameter

A social network is considered compact if it doesn't take
many hops or connections to get from one node to the furthest
away node. The diameter of the network is the distance between the 
two most distant nodes. The following code is a little complicated,
it will be explained later. Note also that this is easier to do
in igraph.

```{r, warning=FALSE,message=FALSE,error=FALSE}
lgc <- component.largest(Moreno,result="graph")
gd <- geodist(lgc)
max(gd$gdist)
```

### Clustering coefficient

True social networks (e.g., made up by people) often exhibit
a tendency towards closure or clustering. One way to assess
this is by measuring the network's transitivity.

```{r}
gtrans(Moreno,mode="graph")
```

In summary, R allows you to get a quick sense of the important
characteristics of any network that is stored as an R network data
object.
