---
title: "Lab 6a - Network Models"
author: "Douglas Luke"
date: "July 16, 2021"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: false
    theme: lumen
    highlight: textmate
urlcolor: blue
---

## Introduction

This lab corresponds to Chapter 10 - Models of Network Structure and Prominence.

We will use `igraph` in this chapter, which has many more functions available to 
create random networks based on mathematical or structural models. Also, some of
the code relies on functions contained in the `tidyverse` suite of packages; namely,
`purrr`.

```{r, message=FALSE}
library(igraph)
library(UserNetR)
library(purrr)
library(lattice)
```



## Erdos-Renyi Random Graphs

These are the simplest models, because they are purely random models. That is, ties between
nodes are determined randomly.

Note that the model functions in igraph have new names, to be more consistent across the set of functions.
The old function names that you find in the book are *deprecated*. That means the old function names
still work, but they may not in future versions of the package.

In `igraph`, there are two functions that can do Erdos-Renyi networks. The only difference is
whether you specify the probability of a tie between two nodes, or how many ties.

```{r}
g <- sample_gnm(n=12,10)
g
graph.density(g)
g <- sample_gnp(n=12,.15)
g
graph.density(g)

```

Run the following a few times to see the random nature of these graphs.

```{r}
op <- par(mar=c(0,1,3,1),mfrow=c(1,2))
plot(sample_gnp(n=25,.15),
                      vertex.color=2,
                      main="First random graph")
plot(sample_gnp(n=25,.15),
                      vertex.color=4,
                      main="Second random graph")
par(op)
```

The shape of the degree distribution is an important characteristic of networks, and helps
to distinguish among different network models. For large random graphs, you should see a
Poisson degree distributions.

```{r}
g <- sample_gnp(n=1000,.005)
plot(degree.distribution(g),
     type="b",xlab="Degree",ylab="Proportion",
     main="Random Graph")
```

E: What happens as the size of the network gets larger? What happens if the probability
of a tie increases or decreases?

The following examples demonstrate some of the emergent characteristics of random
networks. See the chapter for more explanation. However, this code also demonstrates how
statistical analysis can be conducted with network models, to better understand the
full implications of these models.

To understand this simple simulation, you should understand the `apply` family of
iterating functions in R (apply,lapply,sapply).

First, create a function that returns the proportion of nodes in the largest
connected component of a random graph relative to the total number of nodes. (If
the entire network is in one component, then the proportion will be 1.)

```{r}
cmpprp <- function(deg){
  cmp_prp <- max(clusters(sample_gnp(n=1000,
                      p=deg/999))$csize)/1000
  cmp_prp
}
```

Now, we create 500 random networks, for each network there will be
an average degree across the nodes from 1 to 8. Using `map_lgl()` we
will calculate the largest component proportion for each of these 500
networks. (Note that this function requires the `purrr` package.)

```{r}
crnd <- runif(500,1,8)
cmp_vec <- map_dbl(crnd,cmpprp)                 
```

Finally, we can plot the component proportion against the
average degree. This will reveal the relationship
between the average degree in a random graph, and how connected
the graph might be.

```{r}
smoothingSpline <- smooth.spline(crnd,cmp_vec,
                                 spar=0.25)
plot(crnd,cmp_vec,col='grey60',
     xlab="Avg. Degree",
     ylab="Largest Component Proportion")
lines(smoothingSpline,lwd=1.5)
```

The basic point of the above code and figure is that with relatively low average degree, 
any random network will be completely connected.

Second, random networks can still exhibit 'small world' properties.

The plot in the following code uses the `lattice` graphics package. Make sure to download
and install it before running.

```{r}
n_vect <- rep(c(50,100,500,1000,5000),each=50)
g_diam <- sapply(n_vect,function(x) 
  diameter(sample_gnp(n=x,p=6/(x-1))))
```

```{r}
bwplot(g_diam ~ factor(n_vect),
       panel=panel.violin,
       xlab="Network Size",
       ylab="Diameter")

```

Q: Reflect on how the above characteristics of random networks are connected to the idea
of '6-Degrees of Separation' and Milgram's small-world study.

## Small-World Model

The first code chunk shows how small-world networks are constructed, and how they look as the
*rewiring* probability increases. If the rewiring probability is high enough, the network
will just become a random graph.

It also shows how sometimes the easy things are hard in R. The `expression()` and `paste()` parts
of the plot command are just so we can include italics in the figure captions.

```{r}
g1 <- sample_smallworld(dim=1, size=30, nei=2, p=0)
g2 <- sample_smallworld(dim=1, size=30, nei=2, p=.05)
g3 <- sample_smallworld(dim=1, size=30, nei=2, p=.20)
g4 <- sample_smallworld(dim=1, size=30, nei=2, p=1)
op <- par(mar=c(2,1,3,1),mfrow=c(2,2))
plot(g1,vertex.label=NA,layout=layout_with_kk,
     main=expression(paste(italic(p)," = 0")))
plot(g2,vertex.label=NA,
     main=expression(paste(italic(p)," = .05")))
plot(g3,vertex.label=NA,
     main=expression(paste(italic(p)," = .20")))
plot(g4,vertex.label=NA,
     main=expression(paste(italic(p)," = 1")))
par(op)

```

The pure lattice structure that we start with implies that the diameter of a network will be relatively high,
because each node is only connected to close neighbors.

```{r}
g100 <- sample_smallworld(dim=1,size=100,nei=2,p=0)
g100
diameter(g100)

```

The following, rather arduous, code shows how the diameter of a network gets rapidly smaller with modest increases in 
the rewiring probability.

```{r}
p_vect <- rep(1:30,each=10)
g_diam <- map_dbl(p_vect,function(x) 
  diameter(sample_smallworld(dim=1, size=100, 
                               nei=2, p=x/200)))
smoothingSpline = smooth.spline(p_vect, g_diam, 
                                spar=0.35)
plot(jitter(p_vect,1),g_diam,col='grey60',
     xlab="Number of Rewired Edges",
     ylab="Diameter")
lines(smoothingSpline,lwd=1.5)

```

## Scale-free (Power Law) Network Model

First, we define a rescale function which is used in the following plots.

```{r}
rescale <- function(node_char,low,high) {
  min_d <- min(node_char)
  max_d <- max(node_char)
  rscl <- ((high - low)*(node_char - min_d))/(max_d - min_d) + low
  rscl
}
```

The new model function name is `sample_pa()` (for preferential attachment).

```{r}
g <- sample_pa(500,directed=FALSE)
V(g)$color <- 'lightblue'
V(g)[degree(g) > 9]$color <- 'red'
node_size <- rescale(node_char=degree(g),
                     low=2,high=10)
plot(g, vertex.label= NA, vertex.size = node_size)

```

The heavy-tailed degree distribution of a scale-free network is easy to see.

```{r}
degree_distribution(g)
```


```{r, warning=FALSE}
op <- par(mfrow=c(1,2))
plot(degree.distribution(g),xlab="Degree",
     ylab="Proportion")
plot(degree.distribution(g),log='xy',
     xlab="Degree",ylab="Proportion")
par(op)

```

By varying some of the parameters of the model, we can get somewhat more realistic networks.

```{r}
g <- sample_pa(500,out.dist=c(.25,.5,.25),
                   directed=FALSE,zero.appeal = 1)
V(g)$color <- 'lightblue'
V(g)[degree(g) > 9]$color <- 'red'
node_size <- rescale(node_char=degree(g),
                     low=2,high=10)
plot(g, vertex.label= NA, vertex.size = node_size)

```

