---
title: "Lab 6 - Affiliation Networks"
author: "Douglas Luke"
date: "July 16, 2021"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: false
    theme: lumen
    highlight: textmate
urlcolor: blue
---

## Introduction

This lab provides an overview of affiliation networks in R. It follows Chapter 9 in the Springer
book very closely. The code relies primarily on `igraph` functionality.

```{r, message=FALSE}
library(UserNetR)
library(igraph)
library(xtable)
```

### Affiliations as 2-mode networks

Affiliation networks often arrive as an N by K *incidence matrix*. For
example, the following dataframe stores data on students (S) who
attend classes (C).

```{r}
C1 <- c(1,1,1,0,0,0)
C2 <- c(0,1,1,1,0,0)
C3 <- c(0,0,1,1,1,0)
C4 <- c(0,0,0,0,1,1)
aff.df <- data.frame(C1,C2,C3,C4)
row.names(aff.df) <- c("S1","S2","S3","S4","S5","S6")
aff.df
```

### Creating affiliation networks from incidence matrices

Igraph can turn an incidence matrix into a 2-mode affiliation network with one line of code.

```{r}
bn <- graph.incidence(aff.df)
```

This is what it looks like:

```{r}
## Plot setup
plt.x <- c(rep(2,6),rep(4,4))
plt.y <- c(7:2,6:3)
lay <- as.matrix(cbind(plt.x,plt.y))

## Affiliation plot example
shapes <- c("circle","square")
colors <- c("blue","red")
plot(bn,vertex.color=colors[V(bn)$type+1],
     vertex.shape=shapes[V(bn)$type+1],
     vertex.size=10,vertex.label.degree=-pi/2,
     vertex.label.dist=1.2,vertex.label.cex=0.9,
     layout=lay)
```


## Affiliation Network Basics

As shown above, an affiliation network is a bipartite graph,
where there are two types of nodes and there are ties only between
the two types.

The *B* and vertex attribute  called *type* are the two clues
that igraph provides that tell you this is a bipartite graph. The *type*
attribute is what igraph uses to distinguish the two types of nodes.

```{r}
bn
V(bn)$type
V(bn)$name
```

Notice the correspondence between the node labels and the vertex type in the above
output. We can also see the underlying incidence matrix in the 2-mode network using
the `get.incidence()` function.

```{r}
get.incidence(bn)

```

### Creating affiliation networks from edge lists

It is also possible to create affiliation networks from edge lists. This will 
often be the case for large affiliation networks.

The following example recreates the above network, but starting from edgelist data.

Start by creating a dataframe.

```{r}
## Affiliation edgelist
el.df <- data.frame(rbind(c("S1","C1"),
             c("S2","C1"),
             c("S2","C2"),
             c("S3","C1"),
             c("S3","C2"),
             c("S3","C3"),
             c("S4","C2"),
             c("S4","C3"),
             c("S5","C3"),
             c("S5","C4"),
             c("S6","C4")))
el.df
```

Then, the following two steps are necessary to create 
the igraph affiliation network. The first step
creates a basic network, the second step
creates the `type` attribute so that igraph
knows it's a bipartite (affiliation) graph.

```{r}
bn2 <- graph.data.frame(el.df,directed=FALSE)
bn2
V(bn2)$type <- V(bn2)$name %in% el.df[,1]
bn2
```

E: The last piece of code is a little tricky. Can you figure
out what is going on?

Q: How can you tell if `bn` and `bn2` are, in fact, the same networks? 

### Plotting 2-mode networks

Typically, use shape (and maybe color) to
distinguish the two node types.

```{r}
## Simple affiliation plot
shapes <- c("circle","square")
colors <- c("blue","red")
plot(bn,vertex.color=colors[V(bn)$type+1],
     vertex.shape=shapes[V(bn)$type+1],
     vertex.size=10,vertex.label.degree=-pi/2,
     vertex.label.dist=1.2,vertex.label.cex=0.9)
```

### Projections (Creating 1-mode networks from 2-mode networks)

Typically will want to look at one or both of the 1-mode
networks that can be derived from the 2-mode network (e.g., the class network or
the student network).

Simple to create the projections, but then make sure you know which one is which.

```{r}
bn.pr <- bipartite.projection(bn)
bn.pr

```

```{r}
## Bipartite projections 2
graph.density(bn.pr$proj1)
bn.student <- bn.pr$proj1
bn.class <- bn.pr$proj2
graph.density(bn.student)
```

Once you have the projections, then you can explore, analyze, and model
each network to you heart's content!

Look at the underlying data:

```{r}
get.adjacency(bn.student,sparse=FALSE,attr="weight")
get.adjacency(bn.class,sparse=FALSE,attr="weight")

```

Plot them.

```{r}
## One-mode projection plots
shapes <- c("circle","square")
colors <- c("blue","red")
op <- par(mfrow=c(1,2))
plot(bn.student,vertex.color="blue",
     vertex.shape="circle",main="Students",
     edge.width=E(bn.student)$weight*2,
     vertex.size=15,vertex.label.degree=-pi/2,
     vertex.label.dist=1.2,vertex.label.cex=1)
plot(bn.class,vertex.color="red",
     vertex.shape="square",main="Classes",
     edge.width=E(bn.student)$weight*2,
     vertex.size=15,vertex.label.degree=-pi/2,
     vertex.label.dist=1.2,vertex.label.cex=1)
par(op)

```

## Extended Example: Hollywood Actors as an Affiliation Network

The following sections do not have a lot of commentary, see the Springer book
for more explanation.

### 2-mode (actor, movie) affiliation network

Make sure you understand the affiliation network structure of the data.

```{r}
## Hollywood setup
data(hwd)
h1 <- hwd
h1
V(h1)$name[1:10]
V(h1)$type[1:10]
V(h1)$IMDBrating[1:10]
V(h1)$name[155:165]
```

Plotting a small subsection of the network.

```{r}
## Hollywood affiliation setup
V(h1)$shape <- ifelse(V(h1)$type==TRUE,
                      "square","circle")
V(h1)$color <- ifelse(V(h1)$type==TRUE,
                      "red","lightblue")

## Hollywood plot 
h2 <- subgraph.edges(h1, E(h1)[inc(V(h1)[name %in% 
  c("The Wolf of Wall Street","Gangs of New York","The Departed")])])
plot(h2,layout=layout_with_kk)

```

Some exploration of the entire two-mode network.

```{r}
## Graph density
graph.density(h1)

## Hollywood descriptives
table(degree(h1,v=V(h1)[type==FALSE]))
mean(degree(h1,v=V(h1)[type==FALSE]))



```

Busy actors.

```{r}
## Hollywood descriptives 2
V(h1)$deg <- degree(h1)
V(h1)[type==FALSE & deg > 4]$name
busy_actor <- data.frame(cbind(
  Actor = V(h1)[type==FALSE & deg > 4]$name,
  Movies = V(h1)[type==FALSE & deg > 4]$deg
))
busy_actor[order(busy_actor$Movies,decreasing=TRUE),]

```

Some R code to examine popularity.

```{r}
## Hollywood descriptives 3
for (i in 161:1365) {
  V(h1)[i]$totrating <- sum(V(h1)[nei(i)]$IMDBrating)    
   }

## Hollywood descriptives 4
max(V(h1)$totrating,na.rm=TRUE)
pop_actor <- data.frame(cbind(
  Actor = V(h1)[type==FALSE & totrating > 40]$name,
  Popularity = V(h1)[type==FALSE & 
                       totrating > 40]$totrating))
pop_actor[order(pop_actor$Popularity,decreasing=TRUE),]


```

Are popular actors more active (or vice versa)?

```{r}
## Hollywood descriptives 5
for (i in 161:1365) {
  V(h1)[i]$avgrating <- mean(V(h1)[nei(i)]$IMDBrating)    
   }
num <- V(h1)[type==FALSE]$deg
avgpop <- V(h1)[type==FALSE]$avgrating
summary(lm(avgpop ~ num))
## Hollywood popularity
scatter.smooth(num,avgpop,col="lightblue",
               ylim=c(2,10),span=.8,
               xlab="Number of Movies",
               ylab="Avg. Popularity")

```

### 1-mode neworks

More information can be gleaned from the 1-mode projections of the 
Hollywood affiliation network.

This produces the actor and movie 1-mode networks.

```{r}
## Hollywood projections
h1.pr <- bipartite.projection(h1)
h1.act <- h1.pr$proj1
h1.mov <- h1.pr$proj2
h1.act
h1.mov

```

Some characteristics of the movie network.

```{r}
## Movie affiliation network
op <- par(mar = rep(0, 4))
plot(h1.mov,vertex.color="red",
     vertex.shape="circle",
     vertex.size=(V(h1.mov)$IMDBrating)-3,
     vertex.label=NA)
par(op)
```

Descriptive statistics.

```{r}
## Movie descriptives
graph.density(h1.mov)
no.clusters(h1.mov)
clusters(h1.mov)$csize
table(E(h1.mov)$weight)
```

Look at the large connected component of the movie network.

```{r}
## Movie component
h2.mov <- induced.subgraph(h1.mov,
          vids=clusters(h1.mov)$membership==1)

## Plot movie component
plot(h2.mov,vertex.color="red",
     edge.width=sqrt(E(h1.mov)$weight),
     vertex.shape="circle",
     vertex.size=(V(h2.mov)$IMDBrating)-3,
     vertex.label=NA)

```

Examine the cohesive subgroups based on the k-cores. The figure only
includes movies that have coreness greater than 4.

```{r}
## Hollyword cores distribution
table(graph.coreness(h2.mov))

## Plot Hollywood cores
h3.mov <- induced.subgraph(h2.mov,
                  vids=graph.coreness(h2.mov)>4)
h3.mov
plot(h3.mov,vertex.color="red",
     vertex.shape="circle",
    edge.width=sqrt(E(h1.mov)$weight),
    vertex.label.cex=0.7,vertex.label.color="darkgreen",
    vertex.label.dist=0.3,
     vertex.size=(V(h3.mov)$IMDBrating)-3)
```



E: Conduct some similar exploration of the actor 1-mode network.

