---
title: "Lab 6b - Statistical Network Models"
author: "Douglas Luke"
date: "Julyl 16, 2021"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: false
    theme: lumen
    highlight: textmate
urlcolor: blue
---

## Introduction

This lab corresponds to Chapter 11 - Statistical Network Models.

The statistical models covered in this chapter and lab are ERGMs: exponential
random graph models. In R these models are available in the `ergm` package which
is included as part of `statnet`. Note that you should install the `latticeExtra`
package if you want to create the MCMC diagnostics plots for the ERGM models.

```{r, message=FALSE}
library(statnet)
library(UserNetR)
library(latticeExtra)
```



## Background

Review the material presented at the beginning of Chapter 11, particularly the description of
the basic ERGM model and the variety of terms that can be included in the model.

The examples in this chapter make use of network data contained in the `TCnetworks` data object.
Unlike most of the other datasets contained in `UserNetR`, this object is not an `igraph` or `statnet`
network object. Instead, it is a list that contains four different network objects.

The following code loads the list object, extracts the list members into the four network objects.

```{r}
data(TCnetworks)
class(TCnetworks)
TCcnt <- TCnetworks$TCcnt
TCcoll <- TCnetworks$TCcoll
TCdiss <- TCnetworks$TCdiss
TCdist <- TCnetworks$TCdist
summary(TCdiss,print.adj=FALSE)
summary(TCcnt,print.adj=FALSE)
```

As the book and data help file reveal, these network data come from an evaluation of the Indiana
tobacco control program. The first three network objects contain organizational contact, collaboration,
and dissemination tie information. The last network object (TCdist), is a little different, and is
used to contain the geographic distances between all 25 of the program members.

E: Using basic R and `statnet` tools, how can you tell that `TCdist` is different from the other networks?

```{r}
components(TCdiss)
gden(TCdiss)
centralization(TCdiss,betweenness,mode='graph')
```

```{r}
deg <- degree(TCdiss,gmode='graph')
lvl <- TCdiss %v% 'agency_lvl'
plot(TCdiss,usearrows=FALSE,displaylabels=TRUE,
      vertex.cex=log(deg),
      vertex.col=lvl+1,
      label.pos=3,label.cex=.7,
      edge.lwd=0.5,edge.col="grey75")
legend("bottomleft",legend=c("Local","State",
                             "National"),
       col=2:4,pch=19,pt.cex=1.5)

```

## Building ERGM models

As in any type of statistical modeling, it is often useful to start with a baseline, null
model. For `ergm` models, this means no covariates, but the model is conditioned on the
observed number of ties.

```{r}
library(ergm)
DSmod0 <- ergm(TCdiss ~ edges,
               control=control.ergm(seed=40))
class(DSmod0)
summary(DSmod0)
```

The model is estimating the log-odds of observing a tie. So, we can take the logistic
transform of the edges estimated parameter to be able to interpret its meaning. In this case
it means the overall probability of observing a tie in the network, which is equivalent to
its density.

```{r}
plogis(coef(DSmod0))
gden(TCdiss)
```

E: Build null ergm models for the contact and collaboration ties as well. Do they have the same
parameter estimates as for the dissemination ties? Does that make sense to you?

### Node-attribute terms

More informative ERGM models are built up by including other covariates of interest. Types of covariates
in `ergm` are called 'terms.' Terms can be node characteristics, node interactions, relational 
characteristics, and higher order network effects (e.g., degree distributions, triangles, etc.) See
the Harris book and Morris (Morris 2008) for more details. Also, to see the internal
help for ergm terms, type `?ergm.terms`.

The following plot suggests that agencies with more experience in tobacco control may be more
likely to be involved in dissemination ties with other agencies.

```{r}
scatter.smooth(TCdiss %v% 'tob_yrs',
               degree(TCdiss,gmode='graph'),
               xlab='Years of Tobacco Experience',
               ylab='Degree')

```

The following model examines the effects of agency experience, and whether the agency is the 
lead agency, on the likelihood of observing a dissemination tie. This model uses the 
nodefactor term (for a binary node covariate) and the nodecov term (for a quantitative node
covariate).

```{r}
DSmod1 <- ergm(TCdiss ~ edges +
              nodefactor('lead_agency') +
              nodecov('tob_yrs'))
summary(DSmod1)

```

The coefficients of the fitted model can be extracted from the model object. These can be used
to calculate the estimated probability of observing ties for particulary network (dyad) configurations.

So, for example, the above output suggests that agencies with more years of experience
in tobacco control will have increased likelihood of being involved in a dissemination tie.

The following code determines the probability of observing a tie between two non-lead
agencies, each with 5 years of TC experience.

```{r}
coef(DSmod1)
p_edg <- coef(DSmod1)[1]
p_yrs <- coef(DSmod1)[3]
plogis(p_edg + 5*p_yrs + 5*p_yrs)
```

Q: If you take the logistic transform of just the edges parameter in this model, what number do you get, and
how do you interpret it?

### Dyadic terms

One of the powerful features of statistical network models is to explore dyadic effects. That is, how
do characteristics of pairs of actors influence the likelihood that those pairs are connected by a tie?
This allows us to explore and test hypotheses about dyadic similarity (homophily) and dissimilarity
(heterophily).

Before building a model with dyadic terms, it is usually helpful to explore the raw observed tie
patterns with respect to some node categorical attribute. (See the help file for classifiction
codes: ?TCnetworks.)

```{r}
mixingmatrix(TCdiss,'agency_lvl')
mixingmatrix(TCdiss,'agency_cat')
```

Here are three different ways to examine the homophily effects of agency level on dissemination.

```{r}
DSmod2a <- ergm(TCdiss ~ edges +
              nodecov('tob_yrs') +
              nodematch('agency_lvl'),
              control=control.ergm(seed=40))
summary(DSmod2a)
```

```{r}
DSmod2b <- ergm(TCdiss ~ edges +
              nodecov('tob_yrs') +
              nodematch('agency_lvl',diff=TRUE),
              control=control.ergm(seed=40))
summary(DSmod2b)

```

```{r}
DSmod2c <- ergm(TCdiss ~ edges +
              nodecov('tob_yrs') +
              nodemix('agency_lvl',base=1),
              control=control.ergm(seed=40))
summary(DSmod2c)
```

Q: Of the three estimated models, which do you prefer? How does AIC help you in model selection?

Q: Using the differential homophily model, what is the probability of two agencies having a dissemination
tie, when both agencies have five years of tobacco experience, and both are local agencies?

### Relational predictors

Although not clearly described in the papers and documentation for `ergm`, these models can
use other network ties as a relational predictor (covariate). So, for example, you could build
a model that test the hypothese that friends in a high school class (friendship tie) are more
likely to lend money to each other (money flow tie), compared to non-friends.

Here, we build a model that looks at whether contact ties and the distance between
agencies are significantly related
to the probability of observing a dissemination tie. The term we use is 'edgecov'. 

```{r}
as.sociomatrix(TCdist,attrname = 'distance')[1:5,1:5]
as.sociomatrix(TCcnt,attrname = 'contact')[1:5,1:5]
DSmod3 <- ergm(TCdiss ~ edges +
              nodecov('tob_yrs') +
              nodematch('agency_lvl',diff=TRUE) +
              edgecov(TCdist,attr='distance') +
              edgecov(TCcnt,attr='contact'), 
              control=control.ergm(seed=40))
summary(DSmod3)

```

### Local structural terms (dyad dependency)

Local structural predictors can be important elements of a network statistical model, for both
substantive reasons, and to improve model stability. (See Harris book and assigned readings
for more details.)

These types of models are more complicated: they require different estimation techniques (Markov Chain
Monte Carlo, MCMC, for example), they can be less stable, and even with modern computers these models
can take a long time to estimate.

```{r}
DSmod4 <- ergm(TCdiss ~ edges + 
              nodecov('tob_yrs') +
              nodematch('agency_lvl',diff=TRUE) +
              edgecov(TCdist,attr='distance') +
              edgecov(TCcnt,attr="contact") +
              gwesp(0.7, fixed=TRUE), 
              control=control.ergm(seed=40))
summary(DSmod4)

```

## Examining fitted ERGM models
### Examine predicted model profiles

Want to look at estimated probabilities for different model configurations (i.e., covariate patterns) to
make sure you understand the effects and how they are interrelated. See my book for what
these probabilities refer to.

```{r}
prd_prob1 <- plogis(-6.34 + 2*1*.098 + 1.51 + 
                      4*1.042 + .858*(.50^4))
prd_prob1
prd_prob2 <- plogis(-6.34 + 2*5*.098 + 
                      1*1.04 + .858*(.50^4))
prd_prob2

```

### Transform raw parameter estimates into odds-ratios

Because an ERGM model is a type of logistic model (probability of an observed tie), the raw
parameter estimates are appropriately interpreted as log-odds. These can be transformed
into odds-ratios by exponentiating the parameters. See Harris, pp. 50-52 for more details.

This code shows how to extract the appropriate coefficient (and standard error) from the model object, and then use
it to calculate the odds ratio and confidence interval for the `tob_yrs` parameter. Note that sometimes
this is easier to do by hand, rather than remembering the code for extraction.

```{r}
coef(DSmod4)
CF_tobyrs <- coef(DSmod4)[[2]]
CF_tobyrs
summary(DSmod4)$asyse
SE_tobyrs <- summary(DSmod4)$asyse[[2]]
SE_tobyrs
```
Now that we have the appropriate values, the calculations for the OR and CI are straightforward.

```{r}
OR_tobyrs <- exp(CF_tobyrs)
OR_tobyrs
LCI <- exp(CF_tobyrs - 1.96*SE_tobyrs)
UCI <- exp(CF_tobyrs + 1.96*SE_tobyrs)
LCI
UCI
```

E: Interpret the odds-ratio and its confidence interval. Calculate the OR and CI for the first agency level homophily term. How do
you interpret those values?

Q: The double bracket notation can be somewhat confusing for extracting components of lists in R. What happens when you use only a single 
set of brackets?

### Examine model fit
It is always important to assess the convergence and fit of the model. Convergence problems will
be indicated by error or warning messages, or models that produce networks that do not look like
the observed network (density of 1 or 0, for example). Model fit tells us more about how the
modeled network is similar to the observed network.

To examine model fit, `ergm` allows us to create a model fit object.

```{r}
DSmod.fit <- gof(DSmod4, 
                 GOF = ~distance + espartners + 
                   degree + triadcensus,
  	             burnin=1e+5, interval = 1e+5)
summary(DSmod.fit)

```

In addition to looking at the summary, very useful gof plots can be produced:

```{r}
op <- par(mfrow=c(2,2))
plot(DSmod.fit,cex.axis=1.6,cex.label=1.6)
par(op)

```

More sophisticated model diagnostics can also be produced, allowing you to look 'under the hood' at
the estimation process. This can be useful to diagnose model problems.

```{r}
mcmc.diagnostics(DSmod4)
```

### Examine simulated networks based on fitted models

Finally, once you have a model that you think is valid and interesting, you can produce
simulated networks based on the model for further exploration.

```{r}
sim4 <- simulate(DSmod4, nsim=1, seed=569)
summary(sim4,print.adj=FALSE)
```

The following code plots both the observed network and a simulated network based on DSmod4, to
see how they compare. (Note that I found a mistake in the code in the book, and corrected
it here.)

```{r}
op <- par(mfrow=c(1,2),mar=c(0,0,2,0))
lvlobs <- TCdiss %v% 'agency_lvl'
plot(TCdiss,usearrows=FALSE,
      vertex.col=lvlobs+1,
      edge.lwd=0.5,edge.col="grey75",
     main="Observed TC network")
lvl4 <- sim4 %v% 'agency_lvl'
plot(sim4,usearrows=FALSE,
      vertex.col=lvl4+1,
      edge.lwd=0.5,edge.col="grey75",
     main="Simulated network - Model 4")
par(op)

```

E: To better understand how well DSmod4 is doing, you might want to simulate some networks from some of the other
models. In particular, compare a simulated network from the null model (DSmod0) to both the observed network
and our final model (DSmod4).

